---
title: "Clinical Feature Classification"
output: html_document
date: "2026-05-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

remove rows with only NAs
```{r }
library(dplyr)
library (nnet)
library(caret)
library(randomForest)
luad_anot_clean_num <- luad_anot_clean %>% 
  select(ajcc_pathologic_stage, paper_Sex, paper_Age.at.diagnosis, paper_Smoking.Status, paper_expression_subtype, paper_Tumor.stage)%>%
  filter(!is.na(paper_Sex))
```
look how many single NAs are left, there are 19 NA values so we remove all rows with NAs
```{r}
colSums(is.na(luad_anot_clean_num))
luad_anot_clean_num<- na.omit(luad_anot_clean_num)
```
prepare expression_subtype data, convert names to valid R variable names
```{r}
luad_anot_clean_num$paper_expression_subtype[luad_anot_clean_num$paper_expression_subtype == "prox.-inflam"] <- "PI"
luad_anot_clean_num$paper_expression_subtype[luad_anot_clean_num$paper_expression_subtype == "prox.-prolif."] <- "PP"
```
convert the categorical values into numerical ones
```{r}
luad_anot_clean_num$paper_expression_subtype <- factor(luad_anot_clean_num$paper_expression_subtype)
luad_anot_clean_num$paper_Sex <- factor(luad_anot_clean_num$paper_Sex) 
luad_anot_clean_num$paper_Tumor.stage <- factor(luad_anot_clean_num$paper_Tumor.stage)
luad_anot_clean_num$paper_Smoking.Status <- factor(luad_anot_clean_num$paper_Smoking.Status)
luad_anot_clean_num$ajcc_pathologic_stage <- factor(luad_anot_clean_num$ajcc_pathologic_stage)
str(luad_anot_clean_num)
```
divide in train and test data 70/30
```{r}
set.seed(123)
train_d <- createDataPartition(luad_anot_clean_num$paper_expression_subtype, p = 0.7, list = FALSE) 
train_data <- luad_anot_clean_num[train_d, ]
test_data <- luad_anot_clean_num[-train_d, ]
```
calculate a multinomal logistic regression (was too complex with all features so I chose 3: Sex, Tumor.stage, Smoking.status)
```{r}
model_logreg1 <- multinom(paper_expression_subtype ~ paper_Sex + paper_Tumor.stage + paper_Smoking.Status, data = train_data)
summary(model_logreg1)
```
choose different features
```{r}
model_logreg2 <- multinom(paper_expression_subtype ~ paper_Sex + paper_Age.at.diagnosis + paper_Smoking.Status, data = luad_anot_clean_num)
summary(model_logreg2)
```
this model with these 3 features makes the most sense to predict expression_subtype (model has the highest accuracy)
```{r}
model_logreg <-multinom(paper_expression_subtype ~ paper_Tumor.stage + paper_Age.at.diagnosis + paper_Smoking.Status, data = luad_anot_clean_num)
summary(model_logreg)
```
predicting with the test data
```{r}
pred_logreg <- predict(model_logreg, newdata = test_data)
confusionMatrix(
  data = pred_logreg,
  reference = test_data$paper_expression_subtype)
```
Random Forest
```{r}
model_randomforest <- randomForest(paper_expression_subtype ~ ., data = train_data, importance = TRUE, ntree = 500)
#check for the most important features 
varImpPlot(model_randomforest, main = "Feature Importance")
#predict with the test data
pred_randomforest <- predict(model_randomforest, test_data)
confusionMatrix(pred_randomforest, test_data$paper_expression_subtype)
```
k-nearest neighbors
first normalize the data (only features not the outcome, expression_subtype is in column 5 so we are excluding it)
```{r}
data_preproc <- preProcess(train_data[, -5], method = c("center", "scale"))
train_norm <- predict(data_preproc, train_data[, -5])
test_norm <- predict(data_preproc, test_data[, -5])

train_norm$paper_expression_subtype<-train_data$paper_expression_subtype
test_norm$paper_expression_subtype<-test_data$paper_expression_subtype
```
train the model
```{r}
set.seed(123)
model_knn <- train(
  paper_expression_subtype ~ .,
  data = train_norm,
  method = "knn",
  tuneGrid = expand.grid(k = 1:10),
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)
```
testing the knn model
```{r}
pred_knn <- predict(model_knn, newdata = test_norm)
confusionMatrix(pred_knn, test_norm$paper_expression_subtype)
```