---
title: "Clinical Feature Classification"
output: html_document
date: "2026-05-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

remove rows with only NAs
```{r }
library(dplyr)
library (nnet)
library(caret)
library(randomForest)
library(pROC)
```

```{r}
luad_anot_clean[luad_anot_clean == "[unknown]"] <- NA
luad_anot_clean[luad_anot_clean == "[Not Available]"] <- NA
```
 we remove all rows with NAs
```{r}
luad_anot_clean_num <- luad_anot_clean %>% 
  select(ajcc_pathologic_stage, paper_Sex, paper_Age.at.diagnosis, paper_Smoking.Status, paper_expression_subtype, paper_Tumor.stage)
colSums(is.na(luad_anot_clean_num))
luad_anot_clean_num<- na.omit(luad_anot_clean_num)
```
prepare expression_subtype data, convert names to valid R variable names
```{r}
luad_anot_clean_num$paper_expression_subtype[luad_anot_clean_num$paper_expression_subtype == "prox.-inflam"] <- "PI"
luad_anot_clean_num$paper_expression_subtype[luad_anot_clean_num$paper_expression_subtype == "prox.-prolif."] <- "PP"
```
convert the categorical values into numerical ones
```{r}
luad_anot_clean_num$paper_expression_subtype <- factor(luad_anot_clean_num$paper_expression_subtype)
luad_anot_clean_num$paper_Sex <- factor(luad_anot_clean_num$paper_Sex) 
luad_anot_clean_num$paper_Tumor.stage <- factor(luad_anot_clean_num$paper_Tumor.stage)
luad_anot_clean_num$paper_Smoking.Status <- factor(luad_anot_clean_num$paper_Smoking.Status)
luad_anot_clean_num$ajcc_pathologic_stage <- factor(luad_anot_clean_num$ajcc_pathologic_stage)
luad_anot_clean_num$paper_Age.at.diagnosis<-as.numeric(as.character(luad_anot_clean_num$paper_Age.at.diagnosis))
str(luad_anot_clean_num)
```
divide in train and test data 70/30
```{r}
set.seed(123)
train_d <- createDataPartition(luad_anot_clean_num$paper_expression_subtype, p = 0.7, list = FALSE) 
train_data <- luad_anot_clean_num[train_d, ]
test_data <- luad_anot_clean_num[-train_d, ]
```
calculate a multinomal logistic regression (was too complex with all features so I chose 3: Sex, Tumor.stage, Smoking.status)
```{r}
model_logreg1 <- multinom(paper_expression_subtype ~ paper_Sex + paper_Tumor.stage + paper_Smoking.Status, data = train_data)
summary(model_logreg1)
```
choose different features
```{r}
model_logreg2 <- multinom(paper_expression_subtype ~ paper_Sex + paper_Age.at.diagnosis + paper_Tumor.stage, data = luad_anot_clean_num)
summary(model_logreg2)
```
this model with these 3 features makes the most sense to predict expression_subtype (model has the highest accuracy)
```{r}
model_logreg <-multinom(paper_expression_subtype ~ paper_Tumor.stage + paper_Age.at.diagnosis + paper_Smoking.Status, data = luad_anot_clean_num)
summary(model_logreg)
```
predicting with the test data
```{r}
pred_logreg <- predict(model_logreg, newdata = test_data)
confusionMatrix(
  data = pred_logreg,
  reference = test_data$paper_expression_subtype)
```
Random Forest
```{r}
model_randomforest <- randomForest(paper_expression_subtype ~ ., data = train_data, importance = TRUE, ntree = 500)
#check for the most important features 
varImpPlot(model_randomforest, main = "Feature Importance")
#predict with the test data
pred_randomforest <- predict(model_randomforest, test_data)
confusionMatrix(pred_randomforest, test_data$paper_expression_subtype)
```
k-nearest neighbors
first normalize the data (only features not the outcome, expression_subtype is in column 5 so we are excluding it)
```{r}
data_preproc <- preProcess(train_data[, -5], method = c("center", "scale"))
train_norm <- predict(data_preproc, train_data[, -5])
test_norm <- predict(data_preproc, test_data[, -5])

train_norm$paper_expression_subtype<-train_data$paper_expression_subtype
test_norm$paper_expression_subtype<-test_data$paper_expression_subtype
```
train the model
```{r}
set.seed(123)
model_knn <- train(
  paper_expression_subtype ~ .,
  data = train_norm,
  method = "knn",
  tuneGrid = expand.grid(k = 1:10),
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)
```
testing the knn model
```{r}
pred_knn <- predict(model_knn, newdata = test_norm)
confusionMatrix(pred_knn, test_norm$paper_expression_subtype)
```
AUC ROC
```{r}
randomforest_prob <- predict(
  model_randomforest,
  newdata = test_data,
  type = "prob")

roc_rf <- multiclass.roc(
  response = test_data$paper_expression_subtype,
  predictor = as.matrix(randomforest_prob))
roc_rf$auc
```

```{r}
logreg_prob <- predict(
  model_logreg,
  newdata = test_data,
  type = "probs")

roc_logreg <- multiclass.roc(
  response = test_data$paper_expression_subtype,
  predictor = as.matrix(logreg_prob))

roc_logreg$auc
```

```{r}
knn_prob <- predict(
  model_knn,
  newdata = test_data,
  type = "prob")

roc_knn <- multiclass.roc(
  response = test_data$paper_expression_subtype,
  predictor = as.matrix(knn_prob))

roc_knn$auc
```  
```{r}
roc_PI_rf <- roc(
  test_data$paper_expression_subtype == "PI",
  randomforest_prob[, "PI"])
roc_PI_logreg<-roc( test_data$paper_expression_subtype == "PI",
  logreg_prob[, "PI"])
roc_PI_knn<-roc( test_data$paper_expression_subtype == "PI",
  knn_prob[, "PI"])

plot(roc_PI_rf, col="red")
lines(roc_PI_logreg, col="green")
lines(roc_PI_knn, col="blue")
```

```{r}
roc_PP_rf <- roc(
  test_data$paper_expression_subtype == "PP",
  randomforest_prob[, "PP"])
roc_PP_logreg<-roc( test_data$paper_expression_subtype == "PP",
  logreg_prob[, "PP"])
roc_PP_knn<-roc( test_data$paper_expression_subtype == "PP",
  knn_prob[, "PP"])

plot(roc_PP_rf, col="red")
lines(roc_PP_logreg, col="green")
lines(roc_PP_knn, col="blue")
```

```{r}
roc_TRU_rf <- roc(
  test_data$paper_expression_subtype == "TRU",
  randomforest_prob[, "TRU"])
roc_TRU_logreg<-roc( test_data$paper_expression_subtype == "TRU",
  logreg_prob[, "TRU"])
roc_TRU_knn<-roc( test_data$paper_expression_subtype == "TRU",
  knn_prob[, "TRU"])

plot(roc_TRU_rf, col="red")
lines(roc_TRU_logreg, col="green")
lines(roc_TRU_knn, col="blue")
```