---
title: "An Analysis of Clinical Variables and Gene Expression Data across Biopsies from Patients with Lung Adenocarcinomata"
author: "Cedric, Hannah, Merlin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, include = F) # All chunks in need of output need to manually be set to echo = T
```

# Introduction



# Loading and Cleaning of Annotation and Expression Data

 we want to clean up our annotation file. To do this we must first take a look at the file.

```{r, echo=T}
annot <- luad_anot_clean
head(annot)
```

We can see here, that there are indeed values missing in a lot of the columns. Looking at the documentation for the data source ({{SOURCE}}), we can also see, that the first three segments = or the first 12 digits = of the barcode encode the patient ID. We thus replace paper_patient with the barcode trimmed to position 12.

```{r}
annot$paper_patient <- strtrim(annot$barcode, 12)
```

Since we only want the barcode and paper columns, we discard anything not matching to that selection. This leaves us with 20 paper columns.

```{r}
annot <- annot[, c("barcode", grep("^paper_", colnames(annot), value = TRUE))]
length(grep("paper", colnames(annot)))
```

As this also makes the 'paper_' prefix superfluous we strip it from the paper columns. Furthermore, we noticed textual NAs, which were entered into the data set as '[unknown]', '[not available]' and '[Not Available]'. We thus strip these and replace them by real NAs. We perform the same operation on the expression data, whilst we're at it. The paper_Tumor.stage column encodes the Ajcc pathological stage. Thus, we replaced it with the ajcc staging.

```{r, echo=T}
NA_flattener <- function(data) {
  for (i in seq_along(data)) {
    data[tolower(data[[i]]) %in% c("[unknown]","[not available]"), i ] <- NA
  }
  return(data)
}

NA_counter <- function(data, per_col=F) {
  s <- apply(data, 2, function(k) sum(is.na(k)))
  if (per_col) return(s) else return (sum(s))
}
```

```{r}
colnames(annot) <- c(colnames(annot)[1], colnames(annot)[-1] %>% substring(7))
annot$Tumor.stage <- luad_anot_clean$ajcc_pathologic_stage
annot <- NA_flattener(annot)
annot %>% NA_counter(per_col=T) # No NAs in barcode
length(annot$barcode) == length(unique(annot$barcode)) # barcodes can function as unique keys 
annot <- column_to_rownames(annot, "barcode") # Don't like tibbles and their zero-name convention nonsense

exp <- luad_exp_clean
exp <- NA_flattener(exp)
exp %>% NA_counter() # No NAs in whole data set
length(exp$gene) == length(unique(exp$gene)) # genes can functions as unique keys
exp <- column_to_rownames(exp, "gene")
```

When looking at the counts of NAs, the expression data are complete, whilst the annotation data have 5623 NAs in total distributed relatively evenly across all columns, with patient and Tumor.stage being the exceptions.

```{r, echo=T}
NA_counter(exp)
NA_counter(annot)
NA_counter(annot, T)
```

From the head-call above, we know that all of the text columns are characters, even when a factor would be the more appropriate data type. We thus go through the cleaned annotation table and factorise, where appropriate.

```{r}
factor_numeriser <- function(annot) {
  for (i in seq_along(annot)) {
    if (!is.logical(annot[,i])) {
      if (length(unique(annot[,i])) <= 10) {
        annot[,i] <- annot[,i] %>% factor()
      } else {
        out <- tryCatch(
          expr = {temp <- annot[,i] %>% {gsub(",", ".", .)} %>% as.numeric; annot[,i] <- temp},
          warning = function(w) {print(i)},
          error = function(e) {print(i)}
        )
      }
    }
  }
  return(annot)
}

annot <- factor_numeriser(annot)

for (i in seq_along(annot)) print(paste0("Class of column ", i, ": ", class(annot[,i])))
```

We know from the documentation of the data ({{SOURCE}}), that the leading numbers of the fourth segment tell us whether the sample is a tumour or not. We use this to encode a new column entitled 'is.tumour' storing that information. Since this segment only has the values of '01' and '11', and since '11' codes for healthy tissue and '01' for tumour tissue, a boolean suffices.

```{r}
annot <- annot %>% mutate(is.tumour = substring(rownames(annot), 14, 15)=="01")
```

With our annotation data (and incidentally also expression data) cleaned up and somewhat usable, we move on to the next step.

# Data Exploration



## Clinical Annotation



## Expression Data



# Data Analysis



## Dimensionality Reduction of Annotation Data



## Differential Expression Analysis



## Expression Correlation amongst Highly Differentially Expressed Genes



# Predictive Modelling



## Based on Expression Data



## Quality and Comparison of Models



## Based on Clinical Annotation



## Integrative Model



# Discussion



# Conclusion

