---
title: "PCA_Luad"
output: html_document
date: "2026-05-12"
---

```{r setup}
library(dplyr)
library(tidyverse)
library(Rtsne)
library(umap)
load("/Users/USER/Downloads/LUAD_data.RData")
```

## R Markdown
The expression data is first presented in a matrix in which the rows are named after the genes.

```{r matrix}
luad_matrix <- luad_exp_clean %>%
  select(-1) %>%
  as.matrix()
luad_matrix <- scale(luad_matrix)
rownames(luad_matrix) <- luad_exp_clean[[1]]
luad_matrix.tibble = as_tibble(luad_matrix)
```

The variance is calculated for each gene and represented as a vector.

```{r Varianz}
luad_matrix.var = luad_matrix.tibble %>%
  rowwise() %>%
  summarize(variance = var(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  pull(variance)
```

```{r Quantile}
q75 <- quantile(luad_matrix.var, 0.75, na.rm = TRUE)

highVarRows <- which(luad_matrix.var >= q75)
```

We can now perform a principle component analysis. We need to transpose the data, because the PCA function is expecting genes as columns and samples as rows. and we only choose the genes with high Variance.

```{r PCA}
pca = luad_matrix.tibble  %>%
  slice(highVarRows)%>%
  t() %>% 
  prcomp(center = FALSE, scale = FALSE) 
```
We calculate the variance of each principal component and the proportion of the overall variance that the components are explaining.
```{r variance of components}
variance = (pca$sdev)^2
prop.variance = variance/sum(variance)
names(prop.variance) = 1:length(prop.variance)
data.frame(proportion = prop.variance, 
           PCs = as.numeric(names(prop.variance)))

pca_with_annot <- as.data.frame(pca$x) %>%
  tibble::rownames_to_column("barcode") %>%
  merge(luad_anot_clean, by = "barcode")
```
First we want to see if the tumor samples are clustered together and away from the normal tissue samples.
```{r differentiate tumor and normal samples}
pca_with_annot <- pca_with_annot %>%
  mutate(
    sample_code = substr(barcode, 14, 15),
    sample_type = case_when(
      sample_code == "01" ~ "Tumor",
      sample_code == "11" ~ "Normal",
      TRUE ~ "Other" ))
ggplot(pca_with_annot,
       aes(x = PC1, y = PC2, colour = sample_type)) +
  geom_point() +
  scale_colour_manual(values = c("blue", "red"))
ggplot(pca_with_annot,
       aes(x = PC1, y = PC3, colour = sample_type)) +
  geom_point() +
  scale_colour_manual(values = c("blue", "red"))
```

```{r }
ggplot(pca_with_annot,
       aes(x = PC1, y = PC2, colour = paper_expression_subtype)) +
  geom_point() +
  scale_colour_manual(values = c("yellow", "red","blue"))
ggplot(pca_with_annot,
       aes(x = PC1, y = PC3, colour = paper_expression_subtype)) +
  geom_point() +
  scale_colour_manual(values = c( "red","blue", "yellow"))
```
```{r }
ggplot(pca_with_annot,
       aes(x = PC1, y = PC2, colour = paper_Smoking.Status)) +
  geom_point() +
  scale_colour_manual(values = c("yellow", "red","blue", "green", "grey"))
ggplot(pca_with_annot,
       aes(x = PC1, y = PC3, colour = paper_Smoking.Status)) +
  geom_point() +
  scale_colour_manual(values = c( "red","blue", "yellow", "green", "grey"))
```

Rtsne t-distributed Stochastic Neighbor Embedding
```{r Rtsne}
tsne_input <- luad_matrix.tibble %>%
  slice(highVarRows) %>%
  as.matrix() %>%
  t()
set.seed(123)
sample_names <- rownames(tsne_input)

tsne <- Rtsne(
  tsne_input,
  perplexity = 30,
  check_duplicates = FALSE
)

tsne_df <- data.frame(
  barcode=sample_names,
  tSNE1 = tsne$Y[,1],
  tSNE2 = tsne$Y[,2]
)
tsne_with_annot <- tsne_df %>%
  left_join(luad_anot_clean, by = "barcode")%>%
  mutate(
    sample_code = substr(barcode, 14, 15),
    sample_type = case_when(
      sample_code == "01" ~ "Tumor",
      sample_code == "11" ~ "Normal",
      TRUE ~ "Other"
    ))
  
ggplot(tsne_with_annot, aes(tSNE1, tSNE2, colour=sample_type)) +
  geom_point()
```

UMAP Uniform Manifold Approximation and Projection

```{r UMAP} 
umap_input <- t(luad_matrix)
luad.umap <- umap(umap_input)
samples <- rownames(umap_input)

umap_df <- data.frame(
  barcode=samples,
  UMAP1 = luad.umap$layout[,1],
  UMAP2 = luad.umap$layout[,2]
)

umap_with_annot <- umap_df %>%
  left_join(luad_anot_clean, by = "barcode")%>%
  mutate(
    sample_code = substr(barcode, 14, 15),
    sample_type = case_when(
      sample_code == "01" ~ "Tumor",
      sample_code == "11" ~ "Normal",
      TRUE ~ "Other"
    ))
ggplot(umap_with_annot, aes(UMAP1, UMAP2, colour=sample_type)) + geom_point()
```
#UMAP on top PCs
#top k PCs extrahieren
```{r UMAPtopPCs}
pca_umap = luad_matrix.tibble  %>%
  t() %>% 
  prcomp(center = FALSE, scale = FALSE) 

pc_10 <- pca_umap$x[, 1:10]
pc_30 <- pca_umap$x[, 1:30]
pc_50 <- pca_umap$x[, 1:50]

umap_10 <- umap(pc_10)

umap_df_10 <- data.frame(
  barcode=rownames(pc_10),
  UMAP1 = umap_10$layout[,1],
  UMAP2 = umap_10$layout[,2])

umap10_with_annot <- umap_df_10 %>%
  left_join(luad_anot_clean, by = "barcode")

ggplot(umap10_with_annot, aes(UMAP1, UMAP2, colour=paper_expression_subtype)) +
  geom_point()

umap_30 <- umap(pc_30)

umap_df_30 <- data.frame(
  barcode=rownames(pc_30),
  UMAP30_1 = umap_30$layout[,1],
  UMAP30_2 = umap_30$layout[,2])

umap30_with_annot <- umap_df_30 %>%
  left_join(luad_anot_clean, by = "barcode")

ggplot(umap30_with_annot, aes(UMAP30_1, UMAP30_2, colour=paper_expression_subtype)) +
  geom_point()

umap_50 <- umap(pc_50)

umap_df_50 <- data.frame(
  barcode=rownames(pc_50),
  UMAP50_1 = umap_50$layout[,1],
  UMAP50_2 = umap_50$layout[,2])

umap50_with_annot <- umap_df_50 %>%
  left_join(luad_anot_clean, by = "barcode")

ggplot(umap50_with_annot, aes(UMAP50_1, UMAP50_2, colour=paper_expression_subtype)) +
  geom_point()
```
With only the 10 top PCs the clusters are more clearly clustered together. More PCs bring more noise to the UMAP plot. 